\documentclass[11pt]{article}
\usepackage[margin=1.1in]{geometry}
\usepackage{amsmath,amssymb,bm}
\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning}
\usepackage[colorlinks=true,linkcolor=blue,citecolor=blue,urlcolor=blue]{hyperref}

\newcommand{\ud}{\,\mathrm{d}}
\newcommand{\vect}[1]{\bm{#1}}
\newcommand{\Rhat}{\hat{\vect{R}}}
\newcommand{\That}{\hat{\vect{T}}}
\newcommand{\Nhat}{\hat{\vect{N}}}

\title{Minimum-Fuel Elliptic-to-GEO Transfer with Lunar Gravity:\\
Problem Formulation, Third-Body Incorporation, and the First Certified Result}
\author{Michael Casey}
\date{July 22, 2026}

\begin{document}
\maketitle

\begin{abstract}
We extend a certified\footnote{``Certified'' throughout means a
solver-level, \emph{first-order} certificate: full IPOPT convergence
(\texttt{Solve\_Succeeded}), post-hoc numerically re-checked collocation
defects, and endpoint/structure gates --- not a proof of global (or even
local) optimality. See \S\ref{sec:certification} for the exact basis and
\S\ref{sec:minimality} for what a stronger certificate would require.}
two-body low-thrust minimum-fuel campaign --- the
Haberkorn--Martinon--Gergaud (HMG) elliptic-orbit-to-GEO benchmark --- to
include the Moon's gravity at circular-restricted three-body (CR3BP)
fidelity --- i.e.\ the ideal circular-restricted model expressed in a
geocentric, non-rotating frame, not the traditional rotating-frame CR3BP
formulation --- while retaining the two-body campaign's winning
formulation: modified equinoctial elements (MEE) with true longitude as
the independent variable and a \emph{free} total longitude span. The Moon
enters as an Earth-centered third-body perturbation (direct plus indirect
terms) added to the Gauss variational equations as a pure acceleration.
The three-body solution is reached by lunar-gain continuation from the
certified two-body solution on the smooth energy objective, followed by an
energy$\to$fuel homotopy at full lunar mass. At the 10\,N rung the certified
result is $m_f = 1377.1545$\,kg against the two-body certified
$1377.1000$\,kg: the Moon \emph{improves} the transfer by $+54.5$\,g
($+0.004\%$) at the baseline lunar phase, in agreement with an a~priori
tidal null model. This note records the optimal control problem, the lunar
incorporation, the continuation strategy, and the certification basis.
\end{abstract}

\section{The benchmark problem}
\label{sec:benchmark}

The underlying problem is the HMG-2004 benchmark \cite{HMG2004}: transfer a
low-thrust spacecraft of initial mass $m_0 = 1500$\,kg from an inclined
elliptic orbit
\[
P^0 = 11\,625~\text{km}, \qquad e^0 = 0.75, \qquad i^0 = 7^\circ,
\qquad \text{(apogee start, } L_0 = \pi\text{)}
\]
to the geostationary orbit ($P_f = 42\,165$\,km, $e_f = 0$, $i_f = 0$),
maximizing final mass at fixed transfer time $t_f = 1.5\,t_f^{\min}(T)$,
where $t_f^{\min}(T)$ is the certified minimum time at thrust level $T$.
The benchmark constants derive from Caillau--Noailles \cite{CN2001}
($I_{sp} \simeq 1995$\,s; we use the campaign's 2000\,s convention). The
two-body campaign certified the full thrust ladder $10 \to 0.1$\,N; at the
10\,N rung considered here, $t_f = 33.3309$\,TU $= 5.290$\,d and the
certified two-body final mass is $1377.1000$\,kg.

The question posed to the three-body campaign is quantitative: \emph{how
much does lunar gravity move the certified two-body answers} --- final
mass, switch structure, and ultimately the ladder laws --- under identical
boundary conditions and time budget?

\section{The optimal control problem}

\subsection{State, control, and dynamics}

We work in modified equinoctial elements with true longitude $L$ as the
independent variable. The state and control are
\[
X = \left[\,P,\ e_x,\ e_y,\ h_x,\ h_y,\ m,\ t\,\right]^{\!\top},
\qquad
U = \left[\,\vect{\beta};\ \delta\,\right],
\quad \lVert \vect{\beta} \rVert = 1,\ \ \delta \in [0,1],
\]
where $(P, e_x, e_y, h_x, h_y)$ are the usual MEE
($h_x = \tan\tfrac{i}{2}\cos\Omega$, $h_y = \tan\tfrac{i}{2}\sin\Omega$),
$m$ is the mass, $t$ is carried as a state, $\vect{\beta}$ is the unit
thrust direction in the local RTN (radial--transverse--normal) frame, and
$\delta$ is the throttle. With canonical units $\mu_E = 1$,
$\mathrm{LU} = 42\,165$\,km (so GEO circular speed is unity), the induced
time unit is $\mathrm{TU} = \sqrt{\mathrm{LU}^3/\mu_{E,\mathrm{km}}}
\approx 13{,}714$\,s (with $\mu_{E,\mathrm{km}} = 398\,600.4418$\,km$^3$/s$^2$,
consistent with $t_f = 33.3309$\,TU $= 5.290$\,d, \S\ref{sec:benchmark});
every dimensional quantity used below --- the lunar acceleration $a_M$
built from $\mu_M$ and $D_{EM}$ (native km, s), the thrust authority
$T_{\max}/m$ and exhaust velocity $c$ (native m, s), and the time state
$t$ itself --- is converted through this single (LU, TU) pair before
entering the canonical equations. The dynamics that result are the Gauss
variational equations driven by the RTN specific force
$\vect{f} = (f_R, f_T, f_N)$, written here as ordinary \emph{time} rates
$\dot X$ --- i.e., before the $\sigma$-domain transcription applied at the
end of this subsection:
\begin{equation}
\label{eq:gauss}
\begin{aligned}
\dot P   &= \frac{2P}{Z}\sqrt{\tfrac{P}{\mu}}\; f_T, \\
\dot e_x &= \sqrt{\tfrac{P}{\mu}}\,\tfrac{1}{Z}
            \bigl( Z \sin L\, f_R + A_1 f_T - e_y (h_x \sin L - h_y \cos L) f_N \bigr), \\
\dot e_y &= \sqrt{\tfrac{P}{\mu}}\,\tfrac{1}{Z}
            \bigl( -Z \cos L\, f_R + A_2 f_T + e_x (h_x \sin L - h_y \cos L) f_N \bigr), \\
\dot h_x &= \tfrac{1}{2}\sqrt{\tfrac{P}{\mu}}\,\tfrac{X_h}{Z} \cos L\, f_N, \qquad
\dot h_y = \tfrac{1}{2}\sqrt{\tfrac{P}{\mu}}\,\tfrac{X_h}{Z} \sin L\, f_N, \\
\dot L   &= \sqrt{\tfrac{\mu}{P^3}}\,Z^2
            + \sqrt{\tfrac{P}{\mu}}\,\tfrac{1}{Z}\,(h_x \sin L - h_y \cos L)\, f_N, \\
\dot m   &= -\tfrac{T_{\max}}{c}\,\delta, \qquad \dot t = 1,
\end{aligned}
\end{equation}
with $Z = 1 + e_x\cos L + e_y \sin L$, $A_1 = e_x + (1{+}Z)\cos L$,
$A_2 = e_y + (1{+}Z)\sin L$, $X_h = 1 + h_x^2 + h_y^2$, and exhaust
velocity $c = I_{sp} g_0$. In the two-body problem the specific force is
thrust alone, $\vect{f} = (T_{\max}/m)\,\delta\vect{\beta}$.

To make $L$ the independent variable, every trajectory quantity is
transcribed on a fixed grid $\sigma \in [0,1]$ with node longitudes
$L_k = \pi + \sigma_k \Delta L$, i.e.\ $L = \pi + \sigma \Delta L$ and
$dL/d\sigma = \Delta L$. By the chain rule, the state (and the time state
$t$) are propagated in $\sigma$ as
\begin{equation}
\label{eq:sigmadomain}
\frac{dX}{d\sigma} = \Delta L\,\frac{\dot X}{\dot L},
\qquad
\frac{dt}{d\sigma} = \frac{\Delta L}{\dot L},
\end{equation}
i.e.\ every rate in \eqref{eq:gauss} --- including $\dot t = 1$ --- is
divided by $\dot L$ and scaled by $\Delta L$ before collocation. This
division is well posed only if $\dot L > 0$ throughout the transfer
(longitude increasing monotonically); that condition is enforced/checked
in the implementation and holds for this GEO-bound, non-librating
geometry.

\subsection{Free longitude span}

The total swept longitude $\Delta L$ is a single scalar \emph{decision
variable} of the problem --- the transfer's revolution count is chosen by
the optimizer, not prescribed. This is the formulation feature that carried
the two-body ladder through two thrust decades (rev counts from
$\sim$7 to $\sim$700), and preserving it was the decisive argument for
incorporating the Moon as a perturbation in this frame rather than moving
to rotating-frame Cartesian coordinates, where a fixed independent-variable
span makes it difficult in practice --- for a fixed mesh and warm start ---
for the optimizer to add or shed a revolution.

\subsection{Fixing $t_f$: two inequivalent transcriptions}

It is worth being precise about what ``fixed transfer time'' means here,
because two formulations that both \emph{fix $t_f$} are not equivalent.
The direct way is to make time (or a regularized time such as a Sundman
variable) the independent variable and fix the mesh span: the horizon of
the transcription itself is $[0, t_f]$. This is the formulation of our
CR3BP GTO campaigns, and its practical cost is a conditioning/mesh
limitation rather than a mathematical impossibility: with the
independent-variable span fixed, the trajectory's revolution count is tied
to the warm start's geometry, and it is difficult in practice for the
optimizer to add or shed a revolution without re-meshing or a fresh warm
start --- this is the ``topology wall'' those campaigns meet under thrust
continuation.

What we do here is different: the independent variable is $L$ with a
\emph{free} span $\Delta L$, physical time is demoted to a state with
$\dot t = 1$, and the time budget enters only as a terminal
\emph{constraint} $t(1) = t_f$ on that state. The same physical quantity
is fixed, but as a boundary condition rather than as the mesh horizon ---
so the optimizer retains the winding freedom while honoring the time
budget exactly. The distinction parallels the MfMax family of indirect
solvers \cite{HMG2004}: their v0 fixes both $t_f$ and the final longitude
$L_f$; their v1 frees $t_f$ (with transversality) at fixed $L_f$; our
formulation is the reciprocal of v1 --- $t_f$ fixed, $L_f$ free --- and it
is this reciprocal choice that lets one solver family sweep the thrust
ladder, where rev counts change by two orders of magnitude, without ever
re-meshing in time.

\subsection{Objective and homotopy}

The physical objective is maximum final mass, equivalently minimum fuel:
\begin{equation}
\min_{\delta(\cdot),\,\vect{\beta}(\cdot)} \int_0^{t_f} \delta \ud t ,
\qquad t(1) = t_f \ \text{pinned}.
\end{equation}
The bang-bang structure of the $L^1$ objective is reached through the
Bertrand--\'Ep\'enoy convex homotopy \cite{BE2002}
\begin{equation}
J_\varepsilon \;=\; \int_0^{t_f}
\bigl[\, \delta \;-\; \varepsilon\, \delta(1-\delta) \,\bigr] \ud t,
\qquad \varepsilon: 1 \to 0,
\end{equation}
for $\varepsilon \in [0,1]$. At $\varepsilon = 1$ the integrand collapses
to $\delta - \delta(1-\delta) = \delta^2$, so $J_1 = \int_0^{t_f}
\delta^2\,\ud t$ --- a \emph{quadratic throttle (energy) surrogate}, which
we refer to below by the shorthand ``energy problem/objective''; at
$\varepsilon = 0$ the integrand is $\delta$ itself, i.e.\ minimum fuel.
Each member is transcribed by
trapezoidal collocation on the $\sigma$ grid (193 segments at 10\,N,
$\approx 25$ nodes per revolution) and solved as an NLP with
CasADi\,+\,IPOPT, warm-started along the homotopy; a step is accepted only
on \texttt{Solve\_Succeeded} with post-hoc numerically re-checked defects
below $10^{-6}$ (in practice $\sim 10^{-15}$).

\section{Incorporating the Moon}

\subsection{Third-body acceleration with the indirect term}

The Moon is added at CR3BP fidelity in the Earth-centered frame ---
concretely, the ideal circular-restricted model expressed in a geocentric,
non-rotating frame, \emph{not} the traditional rotating-frame CR3BP
formulation or a real ephemeris model (see \S\ref{sec:lunareph} below).
Because that frame is non-inertial --- the Moon accelerates the Earth --- the
perturbing acceleration must carry both the direct and the
\emph{indirect} term:
\begin{equation}
\label{eq:thirdbody}
\vect{a}_M(\vect{r}, t) \;=\; \mu_M \left[
\frac{\vect{r}_M(t) - \vect{r}}{\lVert \vect{r}_M(t) - \vect{r} \rVert^{3}}
\;-\;
\frac{\vect{r}_M(t)}{\lVert \vect{r}_M(t) \rVert^{3}}
\right],
\end{equation}
with $\mu_M/\mu_E = 0.012300$ ($\mu_M = 4902.800$\,km$^3$/s$^2$),
consistent with the CR3BP mass parameter
$\mu^\ast = 0.0121506$ via $\mu^\ast/(1-\mu^\ast)$. Omitting the indirect
term is the classic third-body modeling error; near GEO the two terms are
comparable in size, and their difference is the \emph{tidal} field that
actually perturbs the transfer.

\subsection{Circular lunar ephemeris}
\label{sec:lunareph}

Faithful to the CR3BP idealization, the Moon moves on a circular orbit
\emph{in the reference (equatorial) plane}:
\begin{equation}
\vect{r}_M(t) = D_{EM}
\begin{bmatrix} \cos(n_M t + \varphi_0)\\ \sin(n_M t + \varphi_0)\\ 0 \end{bmatrix},
\qquad
n_M = \sqrt{\frac{\mu_E + \mu_M}{D_{EM}^{3}}},
\end{equation}
with $D_{EM} = 384\,400$\,km ($= 9.1166$\,LU) and sidereal period
$2\pi/n_M = 27.28$\,d. The transfer's $7^\circ$ initial inclination is
measured relative to this plane; the real Moon--equator obliquity
($\sim$18--29$^\circ$) is deliberately not modeled at this stage. The lunar
phase at departure, $\varphi_0$, is a first-class parameter recorded in
every cached artifact; the baseline uses $\varphi_0 = 0$. Because $t$ is
already a state of the L-domain formulation, the time-dependent ephemeris
requires no structural change to the transcription --- the dynamics simply
read the $t$ state.

\subsection{Entry into the Gauss equations}

At each node the Cartesian position is reconstructed from the MEE in closed
form,
\[
\vect{r} = \frac{P/Z}{X_h}
\begin{bmatrix}
(1+\alpha^2)\cos L + 2 h_x h_y \sin L\\
(1-\alpha^2)\sin L + 2 h_x h_y \cos L\\
2\,(h_x \sin L - h_y \cos L)
\end{bmatrix},
\qquad \alpha^2 = h_x^2 - h_y^2,
\]
along with the RTN basis: $\Rhat = \vect{r}/r$, the orbit normal
$\Nhat = [\,2h_y,\ -2h_x,\ 1 - h_x^2 - h_y^2\,]^{\top}/X_h$ (unit and
orthogonal to $\Rhat$ identically), and $\That = \Nhat \times \Rhat$. The
lunar acceleration \eqref{eq:thirdbody} is projected onto this triad and
added to the thrust in the total specific force of \eqref{eq:gauss}:
\begin{equation}
f_R = \tfrac{T_{\max}}{m}\,\delta\beta_R + g\,(\Rhat\!\cdot\!\vect{a}_M),
\qquad
f_T = \tfrac{T_{\max}}{m}\,\delta\beta_T + g\,(\That\!\cdot\!\vect{a}_M),
\qquad
f_N = \tfrac{T_{\max}}{m}\,\delta\beta_N + g\,(\Nhat\!\cdot\!\vect{a}_M),
\end{equation}
where $g \in [0,1]$ is the lunar-gain continuation parameter, multiplying
the lunar force (\S\ref{sec:continuation}). Two coupling rules are essential and are
enforced explicitly in the implementation: the perturbation enters
\emph{every} Gauss equation including the $\dot L$ rate, and it contributes
\emph{nothing} to $\dot m$ --- gravity costs no propellant.

Implementation note: the perturbation is a strictly opt-in branch of the
shared dynamics routine. When absent (or at $g = 0$) the code executes the
\emph{character-identical} original two-body equations, so the certified
two-body campaign is bitwise untouched --- a property enforced by
regression gates rather than assumed. The correctness of the new frame
mathematics is pinned by committed analytic oracles: a radial oracle
(circular equatorial state at $L = 0$ with the Moon on the radial axis,
for which only $\dot e_y$ responds, by exactly $-\sqrt{P/\mu}\,a_R$ in the
time domain of Eq.~\eqref{eq:gauss} --- the coefficient rescales, like
every other rate, under the $\sigma$-domain division \eqref{eq:sigmadomain}),
a transverse oracle at $L = \pi/2$, and a numerical check of $\Nhat$
against the finite-difference orbit normal at random three-dimensional
states.

\section{Solution strategy}
\label{sec:continuation}

\subsection{The pipeline, and the role of the min-time solution}

The full numerical chain behind the certified result is:
\begin{enumerate}
\item \textbf{Min-time (two-body, inherited).} The minimum-time problem is
  \emph{not} re-solved in this campaign. The two-body campaign's certified
  min-time anchor $t_f^{\min}(10\,\mathrm{N})$ enters only through the
  time budget $t_f = 1.5\,t_f^{\min}$ --- the standard fixed-time
  convention that makes the fuel problem well-posed. No min-time
  \emph{trajectory} is used as a seed anywhere in the chain.
\item \textbf{Energy seed (two-body).} A constant-throttle,
  tangential-steering trajectory is propagated from the initial orbit and
  sampled on the $\sigma$ grid (the campaign's two-pass seed: a cheap
  50-node probe fixes the revolution count, then a full-density sample at
  $\approx 25$ nodes/rev). From this seed the smooth two-body
  \emph{energy} problem ($\varepsilon = 1$, no Moon) is solved to full
  convergence. This solve doubles as the certified-path regression gate.
\item \textbf{Lunar-gain continuation (gravity bridge).} The lunar gain is
  walked $g: 0 \to 1$ on the energy objective, warm-starting each step
  (\S\ref{sec:mubridge}).
\item \textbf{$\varepsilon$-homotopy (sharpening).} At full lunar mass the
  energy solution is continued $\varepsilon: 1 \to 0$ down to bang-bang
  minimum fuel (\S\ref{sec:sharpen}).
\end{enumerate}
So the answer to ``min-time first, then min-energy, then fuel?'' is:
\emph{min-time only as the inherited time anchor; energy solved from a
propagated seed, not from a min-time trajectory; and two separate
continuations --- first in the lunar mass at fixed $\varepsilon = 1$, then
in $\varepsilon$ at fixed $g = 1$.} The ordering of the two continuations
is deliberate and is the campaign's central methodological choice
(\S\ref{sec:frompaper}): the physics is changed on the smooth objective,
and the nonsmooth structure is resolved exactly once, at the final physics.

\begin{figure}[htb]
\centering
\begin{tikzpicture}[
  node distance=5.5mm,
  box/.style={rectangle, rounded corners=2pt, draw, fill=blue!7,
              text width=0.56\textwidth, align=center, inner sep=5pt, font=\small},
  gate/.style={font=\scriptsize\itshape, text width=0.32\textwidth, align=left},
  arr/.style={-{Latex[length=2mm]}, thick}]
\node[box] (anchor) {\textbf{Anchors (inherited)}\\
  certified two-body $t_f^{\min}(T)$ $\Rightarrow$ $t_f = 1.5\,t_f^{\min}$};
\node[box, below=of anchor] (seed) {\textbf{Two-pass recipe seed}\\
  tangential constant-throttle propagation; probe fixes $n_{\mathrm{rev}}$,
  full-density sample $N = \mathrm{npr}\cdot n_{\mathrm{rev}}$};
\node[box, below=of seed] (e2b) {\textbf{Two-body energy solve}\\
  $\varepsilon = 1$, $g = 0$ (also the certified-path regression gate)};
\node[box, below=of e2b] (mu) {\textbf{Lunar-gain continuation (gravity bridge)}\\
  $g: 0 \to 0.25 \to 0.5 \to 0.75 \to 1$ on the energy objective,
  warm-started; adaptive bisection on failure};
\node[box, below=of mu] (eps) {\textbf{$\varepsilon$-homotopy (sharpening)}\\
  $\varepsilon: 1 \to 0$ at $g = 1$ (Bertrand--\'Ep\'enoy), warm-started
  per step};
\node[box, below=of eps, fill=green!10] (out) {\textbf{Certified minimum-fuel solution}\\
  states, controls, defect duals, fingerprint, plots, movie};
\draw[arr] (anchor) -- (seed);
\draw[arr] (seed) -- (e2b);
\draw[arr] (e2b) -- (mu);
\draw[arr] (mu) -- (eps);
\draw[arr] (eps) -- (out);
\node[gate, right=3mm of e2b.east] {gate: \texttt{Solve\_Succeeded} $\wedge$
  defect $<10^{-6}$ $\wedge$ unit/terminal errors $<10^{-8}$};
\node[gate, right=3mm of mu.east] {accepted steps only are checkpointed
  (fingerprinted, fail-closed resume)};
\node[gate, right=3mm of eps.east] {per-step caches; endpoint
  $\varepsilon = 0$ required for certification};
\end{tikzpicture}
\caption{The Phase-1 solution pipeline. Every arrow is gated; every accepted
step is checkpointed under a configuration fingerprint, so any stage can
resume without recomputation and no stale configuration can silently
warm-start another.}
\label{fig:pipeline}
\end{figure}

\subsection{What we drew from Bonnard--Caillau--Picot}
\label{sec:frompaper}

Four specific things were taken from \cite{BCP2010}:
\begin{itemize}
\item \textbf{The lunar-gain continuation, in the spirit of mass-ratio
  continuation.} Their \S 4.4 computes Earth--Moon transfers by
  continuation in the mass ratio from the Kepler problem, traced to
  Poincar\'e. Our gravity bridge follows that lineage \emph{in spirit}
  rather than literally: the gain $g$ scales only the lunar force in
  Eq.~\eqref{eq:thirdbody}'s contribution to \eqref{eq:gauss}, while the
  lunar mean motion $n_M$ keeps its full-mass CR3BP value throughout, so
  intermediate steps $g \in (0,1)$ trace a homotopy path, not a physical
  CR3BP solution at a reduced mass ratio --- only the $g = 1$ endpoint is
  exact. Their Proposition~2.3 supplies convergence theory relevant to
  this kind of bridge: conjugate-point clearance is a \emph{sufficient}
  nondegeneracy condition for a smooth continuation branch, not an
  if-and-only-if characterization in this generality (regularity of the
  shooting map and endpoint-rank conditions matter too).
\item \textbf{The benign-landscape rationale for bridging on energy.}
  Their Theorem~3.5 proves the \emph{averaged} Kepler energy problem is a
  Riemannian geodesic flow with \emph{empty conjugate and cut loci} in the
  elliptic domain $e \in [0,1)$. That is a theorem about the
  reduced/averaged Kepler problem, not about our unaveraged,
  three-dimensional, time-dependent MEE problem with free $\Delta L$ and
  collocation; we read it as a strong analogy for why energy-first
  continuation is empirically benign here, not a theorem that certifies
  our landscape. It is nonetheless the best theoretical account on hand
  for why every campaign in this repository has found the energy problem
  the reliable continuation backbone, and the reason the Moon is dialed in
  at $\varepsilon = 1$ rather than on the bang-bang fuel problem.
\item \textbf{Conjugate-point certification (adopted as follow-up).} They
  verify local optimality, in their averaged-energy setting and its own
  regularity hypotheses, by integrating the Jacobi (variational) equation
  and checking the first conjugate time exceeds $t_f$. We attribute this
  as the method used in their setting, not a standalone universal
  certificate. Our indirect-side verification is currently first-order
  only; the Jacobi-field check is the recorded upgrade path
  (\S\ref{sec:secondorder}).
\item \textbf{The $\sigma$-bound rev-count law.} Their
  Propositions~3.4/3.7 bound the control sup-norm by an explicit function
  of the geometry over the cumulated longitude, under their stated
  hypotheses. We treat this as a \emph{prospective comparison} --- to be
  checked against those propositions' exact variables and hypotheses once
  the deep-rung ladder results are in hand, not an inherited law for this
  throttle-limited, fixed-time, free-winding formulation --- the rigorous
  cousin of the campaign's empirical
  $T \cdot t_f^{\min} \approx \mathrm{const}$ law.
\end{itemize}
What we did \emph{not} adopt: their averaging-based analytic initialization
(our discrete two-body solutions are stronger seeds), and their
rotating-frame Cartesian formulation --- correct for their Moon-bound
targets, but our GEO target is Earth-bound throughout, which is exactly the
regime where the MEE/perturbation formulation keeps its free-$\Delta L$
advantage.

\subsection{The two continuations in detail}

The three-body minimum-fuel solution is reached by two continuations run in
sequence, each warm-started and gated:

\paragraph{1. Lunar-gain continuation (gravity bridge).}\label{sec:mubridge}
Starting from the certified two-body \emph{energy} solution
($\varepsilon = 1$), the lunar gain is walked $g: 0 \to 1$
(schedule $0.25, 0.5, 0.75, 1.0$, with adaptive bisection on failure). The
gain $g$ scales only the lunar \emph{force} contribution to
\eqref{eq:gauss}; the lunar mean motion $n_M$ retains its full-mass CR3BP
value throughout, so intermediate steps $g \in (0,1)$ trace a homotopy
path rather than a physical CR3BP solution at reduced mass ratio --- only
the $g = 1$ endpoint is exact. This is done in the spirit of the classical
continuation in the mass ratio traced to Poincar\'e and developed for
low-thrust transfers by Bonnard--Caillau--Picot \cite{BCP2010}, whose
convergence theory ties smoothness of that (literal mass-ratio)
continuation branch to conjugate-point clearance as a sufficient
condition. The continuation is run on the energy objective deliberately:
BCP's Theorem~3.5 shows the \emph{averaged} energy problem is a Riemannian
geodesic flow free of conjugate and cut points in the elliptic domain
\cite{BCP2010} --- a strong analogy, not a proof, for our unaveraged
setting, but the best account on hand for why the energy landscape is
benign here too.

\paragraph{2. $\varepsilon$-homotopy (sharpening) at full Moon.}\label{sec:sharpen}
With $g = 1$ frozen, the standard energy$\to$fuel homotopy
$\varepsilon: 1 \to 0$ is run with the perturbation active, reusing the
two-body campaign's homotopy machinery unchanged. The bang-bang structure
is thus resolved exactly once, at the final physics.

Both stages checkpoint only accepted steps under a dedicated tag namespace
with configuration fingerprints (thrust, mass, $I_{sp}$, $t_f$, $\mu_M$,
$D_{EM}$, $n_M$, $\varphi_0$, $g$), so two-body and three-body caches can
never silently cross-seed.

\section{A priori null model}

Before any solve, the expected size of the lunar effect is scaled by the
tidal acceleration at GEO radius,
$a_{\text{tide}} \approx 2\mu_M r / D_{EM}^3 \approx
\SI{7.3e-6}{\meter\per\second\squared}$, compared with the thrust
authority $T/m_0$:

\begin{center}
\begin{tabular}{cccc}
\toprule
rung & $T/m_0$ [\si{\meter\per\second\squared}] & tide/authority & $t_f$ [lunar months]\\
\midrule
10\,N  & $6.7\times 10^{-3}$ & $0.109\%$ & $0.19$\\
1\,N   & $6.7\times 10^{-4}$ & $1.09\%$  & $\sim 2$\\
0.1\,N & $6.7\times 10^{-5}$ & $10.9\%$  & $\sim 19$\\
\bottomrule
\end{tabular}
\end{center}

\noindent
At 10\,N the Moon is a $\sim 0.1\%$ actor over a fifth of a lunar month:
this is an order-of-magnitude scaling, not a rigorous bound, and it
suggests a sub-$0.1$\,kg mass shift whose sign depends on geometry. The
deep rungs are where the Moon becomes a $\sim 10\%$ actor
acting over many lunar months --- the physically interesting regime, and
the recorded next step.

\section{Results at the 10 N rung}

\paragraph{Bridge (energy objective).}
Gate 1 (perturbation absent; certified-path regression):
\texttt{Solve\_Succeeded}, defect $4.6\times 10^{-15}$, terminal error
$\sim 10^{-36}$. Gate 2 ($g = 0$ with the perturbation wired in): solution
drift $8.0\times 10^{-9}$ relative to gate 1 --- the hook itself introduces
no physics. The gain walk then converged on the \emph{first attempt at
every step}, with defects $4$--$6\times 10^{-15}$ and energy-stage final
mass moving from $1359.68$ to $1359.72$\,kg across the walk --- a
$+40$\,g effect at the energy stage, consistent with the null model.

\paragraph{Certified minimum-fuel result.}
The $\varepsilon$-homotopy reached $\varepsilon = 0$ with all
certification metrics green (defect $4.2\times 10^{-15}$, unit-norm
violation $<10^{-8}$, terminal error $<10^{-8}$, no unintended active
state/control bounds beyond the expected throttle bounds
$\delta \in \{0,1\}$):
\begin{center}
\begin{tabular}{lcccc}
\toprule
 & $m_f$ [kg] & $\Delta m_f$ & switches$^{\ast}$ & $t_f$\\
\midrule
two-body (certified)   & $1377.1000$ & ---        & 19 & $5.290$\,d\\
three-body (this work) & $1377.1545$ & $+0.0545$\,kg ($+0.004\%$) & 19 & $5.290$\,d\\
\bottomrule
\end{tabular}
\end{center}
{\footnotesize $^{\ast}$Nodal counts; switch counts are mesh-band
quantities, not exact integers (campaign mesh-convergence protocol).}

\medskip
At the baseline lunar phase the Moon \emph{helps}: the transfer finishes
$54.5$\,g heavier. The effect sits squarely inside the null-model band,
and the bang-bang structure is unchanged at this rung. Whether the sign of
$\Delta m_f$ is generic or specific to $\varphi_0 = 0$ is an open,
recorded question --- at 10\,N the transfer spans only a fifth of a lunar
month, so the Moon is nearly static during the transfer and the effect
plausibly reverses with phase.

\section{Certification basis and limitations}
\label{sec:certification}

The three-body result is certified on solver metrics: full IPOPT
convergence (\texttt{Solve\_Succeeded}, never
\texttt{Solved\_To\_Acceptable\_Level}), post-hoc numerically re-checked
collocation defects at $10^{-15}$, terminal and unit-norm errors below
$10^{-8}$, homotopy endpoint $\varepsilon = 0$ reached, and no unintended
active state/control bounds beyond the expected throttle bounds. This is a
deliberately weaker bar than the two-body campaign's, which additionally
verifies Pontryagin consistency through a primer-vector reconstruction. A
CR3BP-aware version of that verifier now exists
(\texttt{mee\_primer\_switch.m}, commit \texttt{4353387}): it subtracts
the zero-throttle ballistic/lunar bracket from the primer extraction and
its switching-function term before comparing against the solver's own
switches, and reduces identically to the pre-existing two-body formulas
when the perturbation is absent. Run against the certified 10\,N and 5\,N
CR3BP fuel solutions, the primer/switch gate still \emph{fails}
($32.4^\circ$/$78.4\%$ at 10\,N; $26.7^\circ$/$76.6\%$ at 5\,N) --- but an
A/B check isolating the lunar-specific coupling term (the
$A_0/\dot L_0$ diagnostic, confirmed $\sim 10^{-5}$) shows removing it
moves these numbers by less than $0.1^\circ$. The dominant cause is the
same pre-existing, eccentricity-correlated raw-dual (KKT-multiplier)
anomaly that already fails the pure two-body certified solutions
identically; it is not a lunar-specific defect. Certification here
therefore still rests on the four NLP metrics above, not on primer
agreement, pending that campaign-wide dual-recovery fix landing; a
conjugate-point (Jacobi-field) second-order check in the style of
\cite{BCP2010} remains a further, separate follow-up
(\S\ref{sec:secondorder}).

Further recorded limitations: the Moon-in-reference-plane idealization
(true CR3BP geometry; real obliquity unmodeled); a single lunar phase
$\varphi_0 = 0$ solved so far; and the deep rungs (1, 0.2, 0.1\,N), where
the null model predicts effects two orders of magnitude larger, not yet
run. The adaptive-bisection path of the gravity bridge was never exercised
at 10\,N (all steps converged first-try) and is untested until a deeper
rung stresses it.

\section{Direct versus indirect methods: a working position}

\emph{This section is deliberately speculative} --- it records campaign
\emph{observations}, not a categorical law, to be revised as evidence
accumulates. The organizing axis we have found most predictive is
\textbf{sensitivity versus combinatorics}: indirect (PMP shooting) methods
have tended to degrade with \emph{arc sensitivity} --- the shooting
Jacobian is a state transition matrix whose norm we have observed to grow
roughly exponentially with revolution count, amplified at each deep
perigee pass, \emph{in the single-shooting formulations tried so far} ---
while direct (collocation) methods degrade with \emph{combinatorial
structure} (bang-bang switch discovery) and mesh resolution. Shooting
sensitivity is problem-, scaling-, and formulation-dependent, and is a
known mitigable failure mode in the wider literature: multiple shooting
and conjugate-point-guarded continuation (as in the exception below) both
reduce it. We have not yet tried multiple shooting on this campaign's deep
rungs; the observation above describes what single shooting has done here,
not a proof that it must.

The resulting division of labor, as observed on this repository's
campaigns to date:
\begin{itemize}
\item \textbf{Minimum time at low thrust (many revolutions): direct.}
  All-burn means no switching structure, so collocation's weakness is
  absent, while the long sensitive arc is single shooting's worst case.
  Every certified min-time anchor in this repository is a direct solve.
\item \textbf{Minimum fuel, structure unknown: direct + homotopy for
  discovery.} The $\varepsilon$-continuation finds the bang-bang structure
  without knowing it a priori. Caveat: switch \emph{counts} from a fixed
  mesh are band quantities, not exact integers.
\item \textbf{Precision and certification once seeded: indirect.} With the
  structure known and costates seeded from a converged direct solution, a
  shooting solve has, in the cases run so far, delivered machine-precision
  switch times and costates at low cost --- its weakness (initialization)
  having been paid by the direct stage. This is not automatic: it requires
  a correct multiplier/costate mapping from the direct solution's KKT
  duals and a conditioning check on the resulting shooting Jacobian before
  the result is trusted; neither is guaranteed by seeding alone. With those
  checks passed, the direct-seeded indirect finish is, in our view, a
  strong marriage.
\item \textbf{The instructive exception} \cite{BCP2010}: indirect shooting
  \emph{does} work at 50--60 revolutions when the objective is the smooth
  energy one, the initializer is an averaged analytic solution, and the
  continuation is conjugate-point-guarded --- i.e., high revolution count
  kills \emph{cold or unguarded single} shooting, and nonsmooth structure
  compounds it, but neither is fatal alone, and standard mitigations
  (multiple shooting, guarded continuation) are available.
\end{itemize}
For this campaign specifically: the Phase-1 pipeline is direct end-to-end;
the indirect counterpart (Phase 2) is planned as a direct-seeded finisher
whose payoff is precision switch times and a second-order certificate
(\S\ref{sec:minimality}), not discovery.

\section{Refinement and stronger optimality evidence}
\label{sec:minimality}

\subsection{Switch-time refinement (PSR): yes, eventually}

Both sibling campaigns carry post-solution refinement (PSR) machinery: the
two-body ladder used PSR rounds at its 0.5\,N rung, and the GTO--tulip
campaign built PMP-steered mesh refinement in which the reconstructed
switching function steers local mesh densification at each switch. The same
step belongs in this campaign's future --- switch times from a uniform
$\sigma$-mesh are only mesh-band accurate, and at deep rungs (hundreds of
switches) the uniform mesh under-resolves switch clusters. The prerequisite
this once gated on --- a lunar-aware switching function --- is now built
(\S\ref{sec:certification}, commit \texttt{4353387}): the primer/switch
reconstruction correctly subtracts the ballistic third-body rate under
lunar gravity. What remains open is \emph{not} lunar-specific: PSR consumes
the same primer/switching reconstruction, and that reconstruction still
disagrees with the solver's own switches by tens of degrees at both
certified CR3BP rungs, because of the campaign-wide, eccentricity-
correlated raw-dual (KKT-multiplier) anomaly that fails the pure two-body
certified solutions identically (\S\ref{sec:certification}). PSR is
therefore blocked in practice on that pre-existing anomaly, not on lunar
awareness; the CR3BP-aware primer and PSR remain one work package, but the
gating item has moved from ``make the switching function lunar-aware'' to
``fix the raw-dual recovery the switching function depends on.''

\subsection{Toward genuine local-minimality certificates}
\label{sec:secondorder}

Everything certified so far is \emph{first-order}: IPOPT's KKT conditions,
machine-tight defects, endpoint and structure checks. A converged extremal
satisfying first-order conditions can still be a saddle or a local
maximizer of the induced problem. The plan for genuinely second-order
evidence has three tiers, in order of readiness:
\begin{enumerate}
\item \textbf{NLP-level SOSC (machinery partially built).} The shared
  solver already exposes \texttt{returnModel} --- the live constraint and
  variable registries assembled precisely for KKT/reduced-Hessian work, with
  a design plan (\texttt{PLAN\_sosc.md}) in the two-body campaign. The
  check: project the Lagrangian Hessian onto the null space of the active
  constraints and verify positive definiteness. This certifies the
  \emph{discretized} problem's minimality and is the cheapest tier to
  finish.
\item \textbf{Bang-bang second-order conditions via the induced
  switching-time problem.} For bang-bang arcs the right continuous-time
  frame (Maurer, Osmolovskii) reduces the OCP to a finite-dimensional
  problem in the switch times; positive definiteness of that induced
  Hessian, together with no additional zeros of the switching function, is
  a genuine sufficient condition \emph{under} the accompanying hypotheses
  --- normality of the extremal, regular (strict, transversal) switching
  with no singular arcs, the relevant endpoint constraint qualification,
  and positivity/coercivity of the induced Hessian on the appropriate
  critical cone. Absent those, positive definiteness plus no extra zeros
  certifies only the topology-fixed, finite-dimensional switching-time
  reduction, not the full OCP. Naturally shared with PSR: both live on the
  arc-parameterized representation.
\item \textbf{Indirect-side conjugate-point test.} Integrate the Jacobi
  (variational) equation along the extremal and verify no conjugate time in
  $(0, t_f]$ \cite{BCP2010} --- not, by itself, a generic classical
  certificate: it requires the associated strengthened Legendre condition
  and endpoint regularity hypotheses from BCP's setting, and it certifies
  \emph{weak} local optimality (optimality against nearby extremals of the
  same structure), which is distinct from \emph{strong} local optimality
  and from global minimality. Planned with the Phase-2 indirect solver, and
  dependent on the same CR3BP-aware costate reconstruction as PSR.
\end{enumerate}
Until these land, the honest statement remains: our solutions are
first-order certified extremals whose optimality is corroborated by
basin-reference checks and cross-formulation agreement, not proven minima.

\begin{thebibliography}{9}

\bibitem{HMG2004}
T.~Haberkorn, P.~Martinon, J.~Gergaud,
``Low-thrust minimum-fuel orbital transfer: a homotopic approach,''
\emph{Journal of Guidance, Control, and Dynamics} 27(6), 2004.

\bibitem{CN2001}
J.-B.~Caillau, J.~Noailles,
``Coplanar control of a satellite around the Earth,''
\emph{ESAIM: Control, Optimisation and Calculus of Variations} 6, 2001.

\bibitem{BE2002}
R.~Bertrand, R.~\'Ep\'enoy,
``New smoothing techniques for solving bang-bang optimal control problems,''
\emph{Optimal Control Applications and Methods} 23(4), 2002.

\bibitem{BCP2010}
B.~Bonnard, J.-B.~Caillau, G.~Picot,
``Geometric and numerical techniques in optimal control of two and
three-body problems,''
\emph{Communications in Information and Systems} 10(4), 239--278, 2010.

\bibitem{Betts2010}
J.~T.~Betts,
\emph{Practical Methods for Optimal Control and Estimation Using Nonlinear
Programming}, 2nd ed., SIAM, 2010.

\end{thebibliography}

\end{document}
